Gradual Soundness: Lessons from Static Python

### 2022-01-15 pj-7.1/paper.tex
typo
pj: appendix, import table

\usepackage[framemethod=TikZ]{mdframed}
\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed of untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
One extra step arises when an untyped class overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

All runtimes for this table came from a 64-bit CentOS Linux system using a build
of \SP{} and Cinder compiled with default settings.
Each benchmark version ran 11 times total: once to warm up the JIT and 10 times to
compute an average runtime (using the \code{time} utility).
Every benchmark version invoked the JIT on the same functions---even the untyped versions.

@@

The 25 GitHub issues that we filed during the formalization effort may be found at the following three links:

@@

\section{Pyre Limitations}

FILL condense the email thread in comment (see source)

% Pyre limitations
% 4 messages
% Ben Greenman <benjaminlgreenman@gmail.com>	Fri, Sep 24, 2021 at 11:18 AM
% To: Carl Meyer <carljm@fb.com>
% Cc: Dino Viehland <dinoviehland@fb.com>, "Lu, Kuang-Chen" <kuang-chen_lu@brown.edu>, Shriram Krishnamurthi <shriram@gmail.com>
% >>> When we “add types” to a function (by adopting its module into Static
% >>> Python, meaning Static Python now cares about the type annotations on that
% >>> function), there may be other code calling that function with types that
% >>> don’t match what it is annotated to receive. This then causes new runtime
% >>> type errors for those calls once the module is made static. So sometimes we
% >>> do have to fix those – we consider that to be fixing pre-existing type
% >>> errors that Pyre missed. We probably see a lot less of that because the
% >>> codebase is already type annotated and statically checked by Pyre, than we
% >>> would if we were adopting Static Python into a previously-not-type-checked
% >>> codebase.
% >>
% >> I'm surprised that you see any errors like this. Is Pyre unsound at
% >> these pre-existing spots?
% >
% > Yes, Pyre is unsound in various ways (as a purely static checker of a
% > gradually typed language must inherently be):
% >
% > 1. Some code just isn’t covered or checked by Pyre, because Pyre’s gradual
% > adoption strategy is to only check functions that have annotations in their
% > signature, or in extreme cases because the “code” that’s calling into Python
% > isn’t Python code at all, but rather Python C extension code or Cython code.
% > So Pyre has no visibility into calls from code it isn’t even checking.
% > 2. Any code with annotations involving the `Any` type (or missing
% > annotations) will have values that Pyre types as `Any`, and Pyre will
% > silently allow you to do anything with those values. (It has no other
% > choice, this is the only way that static-only gradual typing can work and
% > actually be gradual.) So anything typed as `Any` by Pyre is a potential
% > unsoundness. And although I say our codebase is largely type annotated,
% > given that lots of it is pre-typing code that can’t be accurately statically
% > annotated without rewriting it, there are plenty of Any annotations
% > (especially `Dict[str, Any]` annotations) floating around.
% > 3. People can (and do) just silence Pyre by adding a `# pyre-fixme` comment
% > that silences a type error on the following line, if they are doing some
% > mass adoption of types or otherwise just don’t feel up to fixing a type
% > error for whatever reason. This tends to happen most often in cases where an
% > annotation is just overly restrictive (e.g. claims to require `str` but in
% > fact the method handles `None` just fine, and the argument really should be
% > annotated as `str | None`.) So we have to fix the annotation to actually
% > match the runtime behavior, and then we also are able to remove the
% > pyre-fixme comments; everyone wins :)
% > 4. If the entire codebase and all its type annotations and type stubs agree
% > to lie about some dynamic wrinkle, Pyre will be satisfied that everything is
% > good (and will have no way to know otherwise). A particular case of this we
% > hit recently is that lots of values annotated as `str` in our codebase can
% > actually at runtime be `Promise` objects that dynamically behave like a
% > `str` if you treat them as one, and implement lazy translation of that
% > string to the current user’s language. So as we adopt such code into Static
% > Python, we have to correct those annotations from `str` to `str | Promise`
% > and handle the fallout. A similar case is the `weakref` module in the Python
% > standard library, which allows you to create “weak references” to objects
% > that will not keep them alive in the garbage collector, but will just
% > resolve to None if the object dies. There’s a `weakref.proxy` that creates a
% > transparent weak-reference proxy to any object, and sometimes methods will
% > return one of those but be annotated as returning the proxied type. This is
% > fine for Pyre, as long as the proxy is sufficiently transparent at the level
% > of Python semantics, but it doesn’t work for Static Python.
% 
% Points 1, 2, and 3 make sense.
% 
% I'm wondering if there are ways to write a fully-typed Pyre program (that
% doesn't use those escape hatches) but is still unsound.* It sounds like
% point 4 might be on this track ... but also that unchecked code (points 1,2)
% might be involved.
% 
% For the weakrefs case, would Pyre approve a function like this one:
% 
% ```
%   def f()->C:
%     return weakref.proxy(C())
% ```
% 
% If so, is that because Pyre comes with a builtin annotation for
% `weakrefs.proxy` that lets it return the Any type?
% 
% 
% * For comparison, TypeScript is unsound by design in a few ways. Any computed
% string can index any object. And all object properties are covariant, despite
% being mutable. I wonder if Pyre has similar "static unsoundness" by design.
% Lu, Kuang-Chen <kuang-chen_lu@brown.edu>	Fri, Sep 24, 2021 at 11:50 AM
% To: Ben Greenman <benjaminlgreenman@gmail.com>
% Cc: Shriram Krishnamurthi <shriram@gmail.com>
% (Internal discussion)
% 
% I think point 4, especially the `str` example, comes from the lack of structural typing. Interfaces can partially solve this problem. But with Nom-style interfaces, they have to define an interface and annotate the classes involved.
% [Quoted text hidden]
% Carl Meyer <carljm@fb.com>	Fri, Sep 24, 2021 at 12:37 PM
% To: Ben Greenman <benjaminlgreenman@gmail.com>
% Cc: Dino Viehland <dinoviehland@fb.com>, "Lu, Kuang-Chen" <kuang-chen_lu@brown.edu>, Shriram Krishnamurthi <shriram@gmail.com>
% [Quoted text hidden]
% I think it could be reasonable to define “fully-typed” as to inherently exclude all four points above! This would also exclude most real-world Python programs, but might still be an interesting definition, because even then there are some unsoundnesses remaining (that I didn’t cover above because they are not usually a problem for us in practice; in most cases I think because they involve parts of the type system we don’t even pretend to support yet in Static Python.) I’ll list some of these known unsoundnesses within the type system below.
% 
% In practice the Python language and standard library include features that are considered too painful to use if fully typed, generally because sound typing would require typing things as `object` (the top type), requiring `isinstance` checks everywhere to keep the static checker happy, so an unsound `Any` is preferred for usability reasons. Weakref proxies are just one example of this; there are many others.
% 
% https://github.com/python/typeshed is the shared repository of type stubs for the Python standard library and popular third-party packages (used by all the static checkers: Pyre, mypy, pyright, pytype…). You can search it for `Any` to find many many more examples.
% 
% 
% > For the weakrefs case, would Pyre approve a function like this one:
% >
% > ```
% >  def f()->C:
% >    return weakref.proxy(C())
% > ```
% >
% > If so, is that because Pyre comes with a builtin annotation for
% > `weakrefs.proxy` that lets it return the Any type?
% 
% Yes, exactly, this is in the shared “typeshed” repository: https://github.com/python/typeshed/blob/master/stdlib/_weakref.pyi#L33
% 
% >
% > * For comparison, TypeScript is unsound by design in a few ways. Any computed
% > string can index any object. And all object properties are covariant, despite
% > being mutable. I wonder if Pyre has similar "static unsoundness" by design.
% 
% Right, and the Python type system has some similar unsound-by-design choices. A few that I’m aware of:
% 
% - Mutable attributes ought to be invariant (because they are in effect both getters and setters, thus appearing in both covariant and contravariant position), thus subclasses should not be allowed to change the type of a base class attribute. But in practice all Python type checkers permit a subclass to narrow the type of a base class attribute. Now that `Final[]` annotations are available on attributes (which would make them soundly covariant), it might be more reasonable to close this hole, here’s discussion on the mypy bug tracker: https://github.com/python/mypy/issues/3208. We are currently sound here in Static Python, but in an odd way; we should error on the subclass attribute type override itself, but instead we ignore it and continue to use the base class attribute type for the subclass also.
% 
% - There are some issues with `typing.Type[]` types (these are the types of class objects themselves) and the fact that they are covariant (which is widely relied on.) This is unsound because incompatible constructor signature overriding is allowed in subclasses (which is also widely used). It’s also unsound because the type of the `self` argument to any overridden method will be Parent on the parent method and Child on the child method, which is the opposite of the normal requirement that method arguments (appearing in a contravariant position) must only be widened on subclass overrides. This means that if some function accepts a type and calls an unbound method on it, providing the `self` arg explicitly, the covariance of `Type` is unsound. On these we are currently OK only because we don’t yet support `Type[]` annotations.
% 
% - Type checking argument splatting in calls is unsound. If we have a dynamic list `l` or dict `d` and we have a call like `foo(*l)` or `foo(**d)`, unless foo accepts `(*args, **kwds)` it is impossible to statically type check the call. Making this sound would have to either require the callee to accept (*args, **kwds) or require that the list or dict is a statically known literal; the latter totally defeats the purpose of argument splatting, the former is more conceivable but still makes it a lot less useful, so in practice type checkers just allow these calls. This one is OK in Static Python because we simply fallback to dynamic calls (and thus runtime argument type checks in the callee) for these calls (I mean it’s OK in that it doesn’t result in cascading unsoundness — it still means runtime error rather than static one for the call itself.)
% 
% There are more cases than just these, but I think the others are more esoteric.
% 
% Carl

@@

FILL

>>>>

\usepackage{lscape}
\usepackage[framemethod=TikZ]{mdframed}
\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
One extra step arises when an untyped class overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

All runtimes for this table came from a 64-bit CentOS Linux system using a build
of \SP{} and Cinder compiled with default settings
(commit \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{\code{d0d071a}}).
Each benchmark version ran 11 times total: once to warm up the JIT and 10 times to
compute an average runtime (using the \code{time} utility).
Every benchmark version invoked the JIT on the same functions---even the untyped versions.

@@

The 25 GitHub issues that arose from our work may be found at the following three links:

@@

\section{On Migrations from Pyre to \SP{}}

Although the Instagram web server has extensive Pyre annotations,
these annotations are not always enough to satisfy \SP{}.
In fact, code may have latent bugs that Pyre did not catch;
we list common reasons for such bugs below:

\begin{itemize}
  \item
    Pyre misses some bugs because it does not monitor run-time interactions
    with untyped code.

  \item
    Similarly, some uses of the Pyre dynamic type (\code{Any}) end up
    raising errors when \SP{} monitors their interactions with precisely-typed code.

  \item
    Occurrences of the special comment \lstinline$# pyre-fixme$ disable Pyre checks.

  \item
    Some annotations intentionally lie to reduce the work of maintaining types.
    For example:

    \begin{itemize}
      \item
        The standard \code{typeshed} repo of Python type annotations declares
        that a weak reference has the dynamic type~\cite{FILL}.
        This declaration is easy to use because it does not force programmers
        to handle the case where their weakly-held object has been collected,
        but it is also a potential source of bugs.
      \item
        Mutable class attributes are covariant in Pyre and other
        Python type checkers, including mypy~\cite{FILL}.
      \item
        The meta-type \code{Type} is unsoundly covariant as well in Pyre
        and others.
      \item
        Argument splatting, i.e. applying a function that expects positional
        arguments to a list (\code{f(*lst)}), is not statically checked
        either.
    \end{itemize}

None of these issues are problems with Pyre.
Quite naturally, it cannot find bugs in code that it does not analyze!
Our point is simply that optional type systems can miss a variety of issues.

@@

\newcommand{\colname}[1]{\textbf{#1}}

The tables in this section present fine-grained data for the microbenchmark programs.
Each benchmark has several versions of its code that accomplish the same work:
the original untyped version (\colname{Orig}) and a typed version (\colname{T-Max}) converted for maximum
performance under Static Python and the JIT.
Some benchmarks have a basic typed version (\colname{T-Min}) that adds only
type annotations and necessary casts to the original code,
without using any more advanced Static Python
features such as primitive types or otherwise optimizing the code for Static
Python.
Fannkuch has a second basic version (\colname{T-Min2}) that uses just a few primitive integers
where it can be done without significant changes to the code.

The other axes of our test matrix are \colname{SP} (whether the Static Python compiler is
used), \colname{JIT} (whether the Cinder JIT is enabled), and \colname{SF} (whether the JIT
shadow-frame mode is enabled, which reduces Python frame allocation costs.) The
matrix is not complete, since using the Static Python compiler on untyped code
has no noticeable effect, and shadow-frame mode is only relevant under the JIT.

Benchmarks are run on a 64-bit CentOS Linux system, using a build of Cinder
compiled with default configure settings. Overall process user time is measured
via the \code{time} utility. Each benchmark/configuration is run 10 times (after an
initial warmup run to ensure there is an up-to-date bytecode cache file and
compilation time will not be included), and all individual data points are
recorded. All data points are in seconds.

JIT runs use a JIT list that includes only the benchmark code itself, to
minimize overhead of compiling unrelated standard library code not used in the
benchmark.

Benchmarks run on the following commit:

\smallskip
{\centering\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}}

Fannkuch in particular is very numeric-heavy, so using unboxed primitive
integers in the typed version can be a big win. For this to work maximally
efficiently, we also need to avoid operations that don't yet support primitives
and thus would require expensive boxing. In typed-and-optimized Fannkuch (\colname{T-Max}), these changes move
a lot of the work out of heavily-optimized C builtin methods (for list
slice/insert/pop) and into simple Python code doing the equivalent operations
manually with simple iteration and indexing. In the short run this is a massive
pessimization for non-static Python (or even SP without the JIT, since
currently we always use boxed integers in the interpreter loop.) In the long
run it suggests that Static Python + JIT could make it feasible to implement
many more of these core datastructure operations in pure Python rather than in
C.

Because of this subtlety, we include benchmark results for both typed and
untyped versions of each benchmark. In the short term, perhaps the most
relevant comparison is the performance of Static Python on the typed benchmark
vs the performance of nonstatic Python on the untyped benchmark; this gives a
fair picture of expected perf gains converting code to Static Python with some
willingness to optimize.

We also observe significant untapped opportunity to improve Static Python's
performance; for example, $20\%$ of Typed Fannkuch SP+JIT time is spent on
bounds-checking array accesses; most of this runtime bounds-checking cost could
be eliminated if the Static Python compiler tracks known array sizes and
integer values more thoroughly. And it would also be possible to support list
slicing, list pop, and list insert with primitive arguments, thus avoiding the
need to rewrite these operations in Python for maximum efficiency.

\colname{T-Min} Fannkuch without the JIT is slow because of a known issue where we
don't cache function lookups for \code{INVOKE\_FUNCTION} in the interpreter loop,
making the calls to \code{list.insert} and \code{list.pop} slower; this is fixable.

\begin{landscape}

\begin{table}[tph]
  \caption{Richards microbenchmark data}
  \label{t:mb:richards}
  \footnotesize

  \begin{tabular}{rrrrrrrrrrrrrrr}
  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{Orig} & \colname{Orig} & \colname{Orig} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $0.57$ & $1.29$ & $7.86$ & $8.36$ & $10.24$ & $16$ & $1.41$ & $2.68$ & $7.67$ & $5.58$ & $7.97$ & $13.32$ & $2.6$ & $3.88$ & $10.17$ \\
  $0.59$ & $1.32$ & $7.92$ & $8.84$ & $9.97$ & $15.73$ & $1.38$ & $2.6$ & $7.9$ & $5.92$ & $7.25$ & $13.59$ & $2.6$ & $4.01$ & $10.18$ \\
  $0.58$ & $1.33$ & $7.89$ & $8.9$ & $10.57$ & $16.2$ & $1.39$ & $2.64$ & $7.74$ & $5.54$ & $7.64$ & $13.83$ & $2.61$ & $4.17$ & $10.73$ \\
  $0.58$ & $1.33$ & $7.92$ & $8.53$ & $10.71$ & $16.71$ & $1.38$ & $2.65$ & $7.8$ & $6.06$ & $7.24$ & $13.95$ & $2.9$ & $3.92$ & $9.96$ \\
  $0.56$ & $1.33$ & $7.83$ & $9.15$ & $9.89$ & $16.62$ & $1.37$ & $2.62$ & $7.68$ & $5.52$ & $7.2$ & $13.5$ & $2.63$ & $3.79$ & $10.04$ \\
  $0.56$ & $1.33$ & $7.97$ & $8.25$ & $10.44$ & $15.65$ & $1.39$ & $2.58$ & $7.8$ & $5.63$ & $7.27$ & $13.52$ & $2.64$ & $3.95$ & $9.99$ \\
  $0.55$ & $1.32$ & $8.18$ & $8.78$ & $10.07$ & $16.22$ & $1.36$ & $2.65$ & $7.79$ & $5.74$ & $7.17$ & $13.34$ & $2.62$ & $3.9$ & $10.16$ \\
  $0.55$ & $1.31$ & $7.85$ & $8.67$ & $10.07$ & $15.73$ & $1.43$ & $2.66$ & $7.82$ & $5.56$ & $7.54$ & $13.68$ & $2.6$ & $3.84$ & $9.69$ \\
  $0.56$ & $1.31$ & $7.96$ & $8.52$ & $10.14$ & $15.82$ & $1.53$ & $2.63$ & $7.9$ & $5.59$ & $6.89$ & $13.79$ & $2.6$ & $4.04$ & $9.82$ \\
  $0.57$ & $1.32$ & $8$ & $8.35$ & $10.57$ & $16.66$ & $1.37$ & $2.71$ & $7.95$ & $5.74$ & $6.91$ & $13.5$ & $2.62$ & $3.85$ & $9.85$ \\
\end{tabular}

\end{table}
\end{landscape}

\begin{landscape}
\begin{table}[tph]
  \caption{Fannkuch microbenchmark data}
  \label{t:mb:fannkuch}
  \footnotesize

  \begin{tabular}{rrrrrrrrrrrrrrr}
  \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Max} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min} & \colname{T-Min2} & \colname{T-Min2} & \colname{T-Min2} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{SP} \colname{JIT} \colname{SF}) & (\colname{SP} \colname{JIT}) & (\colname{SP}) \\\hline
  $1.27$ & $1.39$ & $34.69$ & $40.74$ & $41$ & $48.23$ & $2.77$ & $2.83$ & $6.18$ & $2.89$ & $2.86$ & $4.07$ & $2.48$ & $2.52$ & $7.28$ \\
  $1.32$ & $1.25$ & $34.35$ & $40.23$ & $40.94$ & $49.03$ & $2.85$ & $2.8$ & $6.12$ & $2.94$ & $2.9$ & $4.06$ & $2.52$ & $2.48$ & $7.28$ \\
  $1.28$ & $1.26$ & $34.67$ & $40.74$ & $40.65$ & $49.42$ & $2.79$ & $2.77$ & $6.38$ & $2.87$ & $2.86$ & $4.1$ & $2.47$ & $2.53$ & $7.24$ \\
  $1.23$ & $1.31$ & $35.71$ & $40.94$ & $39.7$ & $49.35$ & $2.87$ & $2.81$ & $6.23$ & $2.87$ & $2.85$ & $4.29$ & $2.48$ & $2.57$ & $7.35$ \\
  $1.28$ & $1.38$ & $34.85$ & $39.59$ & $40.46$ & $49.18$ & $2.85$ & $2.84$ & $6.17$ & $2.92$ & $2.81$ & $4.14$ & $2.5$ & $2.56$ & $7.34$ \\
  $1.26$ & $1.34$ & $33.96$ & $40.45$ & $39.99$ & $47.79$ & $2.87$ & $2.8$ & $6.21$ & $2.94$ & $2.8$ & $4.06$ & $2.43$ & $2.47$ & $7.17$ \\
  $1.26$ & $1.23$ & $34.65$ & $40.78$ & $40.97$ & $49.7$ & $2.92$ & $2.76$ & $6.04$ & $3.11$ & $3.05$ & $4.08$ & $2.47$ & $2.69$ & $7.18$ \\
  $1.27$ & $1.35$ & $33.83$ & $40.96$ & $41.78$ & $47.92$ & $2.78$ & $2.76$ & $6.13$ & $3.1$ & $2.86$ & $4.19$ & $2.47$ & $2.45$ & $7.14$ \\
  $1.23$ & $1.4$ & $36.23$ & $40.49$ & $40.29$ & $48.68$ & $2.84$ & $2.76$ & $6.13$ & $2.87$ & $2.96$ & $4.31$ & $2.43$ & $2.55$ & $7.23$ \\
  $1.23$ & $1.27$ & $36.05$ & $40.72$ & $40.11$ & $48.29$ & $2.92$ & $2.79$ & $6.17$ & $2.91$ & $2.96$ & $4.1$ & $2.52$ & $2.49$ & $7.2$ \\
\end{tabular}

\begin{tabular}{rrrrrr}
  \colname{T-Min2} & \colname{T-Min2} & \colname{T-Min2} & \colname{Orig} & \colname{Orig} & \colname{Orig} \\
  (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () & (\colname{JIT} \colname{SF}) & (\colname{JIT}) & () \\\hline
  $3.16$ & $3.08$ & $4.59$ & $2.81$ & $2.94$ & $4.02$ \\
  $3.26$ & $3.08$ & $4.51$ & $2.66$ & $2.85$ & $4.22$ \\
  $3.11$ & $3.18$ & $4.59$ & $2.69$ & $2.78$ & $4.01$ \\
  $3.18$ & $3.14$ & $4.73$ & $2.75$ & $2.8$ & $4.01$ \\
  $3.14$ & $3.12$ & $4.63$ & $2.75$ & $2.92$ & $4.07$ \\
  $3.23$ & $3.09$ & $4.53$ & $2.78$ & $2.83$ & $4.36$ \\
  $3.41$ & $3.21$ & $4.48$ & $2.84$ & $2.81$ & $4.19$ \\
  $3.43$ & $3.14$ & $4.53$ & $2.7$ & $2.86$ & $4.25$ \\
  $3.08$ & $3.24$ & $4.53$ & $2.76$ & $2.78$ & $4.08$ \\
  $3.16$ & $3.18$ & $4.45$ & $2.89$ & $2.76$ & $4.16$ \\
\end{tabular}
\end{table}

\end{landscape}

\begin{table}[tph]
  \caption{Exact commands to invoke each benchmark}
  \label{t:mb:commands}
  \tiny

  \begin{tabular}{ll}
  \colname{Benchmark} & \colname{Command} \\\hline
  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
  Richards \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static.py 100} \\
  (\colname{SP}) &  \\
  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/richards\_static.py 100} \\
  Richards \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{JIT}) & \texttt{ Tools/benchmarks/richards\_static.py 100} \\
  Richards \colname{T-Max} & \texttt{time ./python Tools/benchmarks/richards\_static.py 100} \\
  () &  \\
  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
  Richards \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/richards\_static\_basic.py 100} \\
  (\colname{SP}) &  \\
  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/richards\_static\_basic.py 100} \\
  Richards \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_richards\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{JIT}) & \texttt{dcards Tools/benchmarks/richards\_static\_basic.py 100} \\
  Richards \colname{T-Min} & \texttt{time ./python Tools/benchmarks/richards\_static\_basic.py 100} \\
  () &  \\
  Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
  (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/richards.py 100} \\
  Richards \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
  (\colname{JIT}) & \texttt{hmarks/richards.py 100} \\
  Richards \colname{Orig} & \texttt{time ./python Tools/benchmarks/richards.py 100} \\
  () &  \\
  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{SP} \colname{JIT}) & \texttt{ -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
  Fannkuch \colname{T-Max} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static.py 5} \\
  (\colname{SP}) &  \\
  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{JIT} \colname{SF}) & \texttt{ -X jit-shadow-frame Tools/benchmarks/fannkuch\_static.py 5} \\
  Fannkuch \colname{T-Max} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static.txt -X jit-enable-jit-list-wildcards} \\
  (\colname{JIT}) & \texttt{ Tools/benchmarks/fannkuch\_static.py 5} \\
  Fannkuch \colname{T-Max} & \texttt{time ./python Tools/benchmarks/fannkuch\_static.py 5} \\
  () &  \\
  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{SP} \colname{JIT}) & \texttt{dcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
  Fannkuch \colname{T-Min} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
  (\colname{SP}) &  \\
  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{JIT} \colname{SF}) & \texttt{dcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
  Fannkuch \colname{T-Min} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic.txt -X jit-enable-jit-list-wil} \\
  (\colname{JIT}) & \texttt{dcards Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
  Fannkuch \colname{T-Min} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic.py 5} \\
  () &  \\
  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
  (\colname{SP} \colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
  (\colname{SP} \colname{JIT}) & \texttt{ldcards -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
  Fannkuch \colname{T-Min2} & \texttt{time ./python -X install-strict-loader Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
  (\colname{SP}) &  \\
  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
  (\colname{JIT} \colname{SF}) & \texttt{ldcards -X jit-shadow-frame Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
  Fannkuch \colname{T-Min2} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_fannkuch\_static\_basic2.txt -X jit-enable-jit-list-wi} \\
  (\colname{JIT}) & \texttt{ldcards Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
  Fannkuch \colname{T-Min2} & \texttt{time ./python Tools/benchmarks/fannkuch\_static\_basic2.py 5} \\
  () &  \\
  Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards -X jit-sha} \\
  (\colname{JIT} \colname{SF}) & \texttt{dow-frame Tools/benchmarks/fannkuch.py 5} \\
  Fannkuch \colname{Orig} & \texttt{time ./python -X jit -X jit-list-file=Tools/benchmarks/jitlist\_main.txt -X jit-enable-jit-list-wildcards Tools/benc} \\
  (\colname{JIT}) & \texttt{hmarks/fannkuch.py 5} \\
